humanCdGF_predict_cfu

### 2016-11-07 code/HumanCdGF_predict_cfu.Rmd
add caret::rfe to col rate

rfe_ctrl <- rfeControl(functions = rfRFE,
                   method = "repeatedcv",
                   repeats = 10,
                   verbose = FALSE)
subsets <- c(1:5, c(seq(10, 70, 10)))
set.seed(seed)
rf_model <- rfe(cfu_shared_colonized.df[ , -1], cfu_shared_colonized.df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
top_important_OTU <- predictors(rf_model)

rf_n_features <- rf_model$fit
rf_prediction <- data.frame(observed = cfu_shared_colonized.df$log_cfu_med, 
                       predicted = rf_n_features$predicted)
rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')

@@

layout(matrix(c(1:16), ncol=4, byrow=TRUE))
for(otu in seq_along(importance_om$tax)){
    partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), x.var=rownames(importance_om)[otu],
                xlab='', main=paste(importance_om$tax_label[otu]))
    par(new = T)
    plot(y = partial_df[ , 'log_cfu_med'],
         x = partial_df[ , rownames(importance_om)[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Log10 CFU')
    if(length(features_cor[[otu]]) > 1){
    mtext(side = 1, line = 3, cex = 0.6,
          paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
                paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ',')))
    }
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(partial_df$cage_id)), function(m){
    mouse <- unique(partial_df$cage_id)[m]
    points(y = partial_df[partial_df$cage_id %in% mouse, 'log_cfu_med'],
           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```

@@

```{r predict colonization rate, fig.width=10,fig.height=10}

meta_data %>% 
  filter(day %in% c(1,2,3,4), !cage_id == 'NP1', 
         Early_Euth == FALSE, cdiff_strain == 431) %>% 
  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
    theme(legend.justification=c(1,0), legend.position=c(1,0), 
          legend.title=element_blank())

@@

col_rate <- sapply(mouse_list, function(i){
  test <- meta_data %>% 
    filter(!is.na(day), mouse_id == i) %>% 
    select(day, log_cfu) %>% 
    filter(!is.na(log_cfu))
  test$rel_cr <- test$log_cfu/median(test$log_cfu[test$day %in% c(4:10)])
  colonization_ratio <- sum(test$rel_cr[test$day %in% c(1:4)])/sum(test$day %in% c(1:4))})
col_rate <- data.frame(mouse_id = names(col_rate), ratio = col_rate, 
                       stringsAsFactors = F, row.names = NULL)

col_rate_apply <- meta_data %>% 
  select(mouse_id, cage_id) %>% 
  unique() %>% 
  right_join(col_rate)

shared_col_rate.df <- col_rate %>% 
    inner_join(community_d0.df)
shared_colonized.df <- select(shared_col_rate.df, -mouse_id, -cage_id)

set.seed(seed)
rf_col_rate <- randomForest(ratio ~ ., data = shared_colonized.df,
                           importance = TRUE, ntree = n_trees, mtry = 14)

importance_sorted <- sort(importance(rf_col_rate)[,1], decreasing = T)
top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))

top_col_rate_df <- data.frame(shared_colonized.df %>% 
  select(ratio, one_of(top_important_OTU)))

set.seed(seed)
rf_col_rate_features <- randomForest(ratio ~ ., data=top_col_rate_df,
                              importance=TRUE, ntree=n_trees, mtry = 2,
                              keep.forest=TRUE)

importance_final <- sort(importance(rf_col_rate_features)[,1], decreasing = T)

rf_col_rate_prediction <- data.frame(observed = shared_colonized.df$ratio, 
                       predicted = rf_col_rate_features$predicted)
rf_cr_prediction <- cbind(cage_id = shared_col_rate.df[,'cage_id'], rf_col_rate_prediction)

ggplot(data = rf_cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
  labs(x= 'Observed', y = 'Predicted',
       title = "Random Forest Prediction of Relative Day 1 Colonization Level of C. Difficile from Day 0 Community") + 
  theme_bw() + xlim(0,1) + ylim(0,1) + 
  annotate('text', x=0.15*(max(rf_cr_prediction$observed)), 
           y=0.95*(max(rf_cr_prediction$predicted)), parse = T,
           label=paste("R^2 == ", round(tail(rf_col_rate_features$rsq, 1), 2))) # + 
#  theme(legend.position = 'none')

importance.df <- data.frame(otu = names(importance_final),
                            importance = importance_final,
                            get_tax(1, names(importance_final), tax_file))

@@

cage_color <- rainbow(length(unique(rf_cr_prediction$cage_id)))
for (i in seq_along(importance_final)) {
    partialPlot(rf_col_rate_features, top_col_rate_df, names(importance_final)[i],
                xlab='', main=paste(importance.df$tax_label[i]))#, log = 'x')
    par(new = T)
    plot(y = top_col_rate_df[ , 'ratio'],
         x = top_col_rate_df[ , names(importance_final)[i]], 
         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Relative Colonization')
# To color points by cage - Change pch to 21 and un-comment following lines
#  for(m in seq_along(unique(rf_cr_prediction$cage_id))){
#    cage <- unique(rf_cr_prediction$cage_id)[m]
#    points(y = top_col_rate_df[rf_cr_prediction$cage_id %in% cage, 'ratio'],
#         x = top_col_rate_df[rf_cr_prediction$cage_id %in% cage , names(importance_final)[i]], 
#         pch = 21, bg = cage_color[m])
#  }
}
```

>>>>

rfe_ctrl <- rfeControl(functions = rfRFE,
                   method = "repeatedcv",
                   repeats = 10,
                   verbose = FALSE)
subsets <- c(1:5, c(seq(10, 70, 10)))
set.seed(seed)
rf_cfu_model <- rfe(cfu_shared_colonized.df[ , -1], cfu_shared_colonized.df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
top_important_OTU <- predictors(rf_cfu_model)

rf_n_features <- rf_cfu_model$fit
rf_prediction <- data.frame(observed = cfu_shared_colonized.df$log_cfu_med, 
                       predicted = rf_n_features$predicted)
rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')

@@

layout(matrix(c(1:9), ncol=3, byrow=TRUE))
for(otu in seq_along(importance_om$tax)){
    partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), x.var=rownames(importance_om)[otu],
                xlab='', main=paste(importance_om$tax_label[otu]))
    par(new = T)
    plot(y = partial_df[ , 'log_cfu_med'],
         x = partial_df[ , rownames(importance_om)[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Log10 CFU')
    if(length(features_cor[[otu]]) > 1){
    mtext(side = 1, line = 3, cex = 0.6,
          paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
                paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ',')))
    }
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(partial_df$cage_id)), function(m){
    mouse <- unique(partial_df$cage_id)[m]
    points(y = partial_df[partial_df$cage_id %in% mouse, 'log_cfu_med'],
           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```

@@

```{r predict colonization rate, fig.width=10,fig.height=10}
# plot initial C. difficile colonization of mice 
#meta_data %>% 
#  filter(day %in% c(1,2,3,4), !cage_id == 'NP1', 
#         Early_Euth == FALSE, cdiff_strain == 431) %>% 
#  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
#    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
#    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
#    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
#   theme(legend.justification=c(1,0), legend.position=c(1,0), 
#          legend.title=element_blank())

@@

col_rate <- sapply(mouse_list, function(i){
  test <- meta_data %>% 
    filter(!is.na(day), mouse_id == i) %>% 
    select(day, log_cfu) %>% 
    filter(!is.na(log_cfu))
  test$rel_cr <- test$log_cfu/median(test$log_cfu[test$day %in% c(4:10)])
  colonization_ratio <- sum(test$rel_cr[test$day %in% c(1:4)])/sum(test$day %in% c(1:4))})
col_rate <- meta_data %>% 
  select(mouse_id, cage_id) %>% 
  unique() %>% 
  right_join(data.frame(mouse_id = names(col_rate), ratio = col_rate, 
                       stringsAsFactors = F, row.names = NULL))

shared_col_rate_df <- col_rate %>% 
    inner_join(community_d0.df)
shared_cr_df <- select(shared_col_rate_df, -mouse_id, -cage_id)

set.seed(seed)
rf_cr_model <- rfe(shared_cr_df[ , -1], shared_cr_df$ratio,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cr_features <- predictors(rf_cr_model)
rf_col_rate <- rf_cr_model$fit

cr_prediction <- data.frame(observed = shared_cr_df$ratio, 
                       predicted = rf_col_rate$predicted)
cr_prediction <- merge(cr_prediction, col_rate, by.x = 'observed', by.y = 'ratio')

#top_col_rate_df <- data.frame(shared_colonized.df %>% 
#  select(ratio, one_of(cr_features)) %>% 
#  left_join(col_rate))

#set.seed(seed)
#rf_col_rate_features <- randomForest(ratio ~ ., data=top_col_rate_df,
#                              importance=TRUE, ntree=n_trees, mtry = 2,
#                              keep.forest=TRUE)
#importance_final <- sort(importance(rf_col_rate_features)[,1], decreasing = T)
#rf_col_rate_prediction <- data.frame(observed = shared_colonized.df$ratio, 
#                       predicted = rf_col_rate_features$predicted)
#rf_cr_prediction <- cbind(cage_id = shared_col_rate.df[,'cage_id'], rf_col_rate_prediction)

ggplot(data = cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
  labs(x= 'Observed', y = 'Predicted',
       title = "Random Forest Prediction of Relative Day 1 Colonization Level of C. Difficile from Day 0 Community") + 
  theme_bw() + xlim(0,1) + ylim(0,1) + 
  annotate('text', x=0.15*(max(cr_prediction$observed)), 
           y=0.95*(max(cr_prediction$predicted)), parse = T,
           label=paste("R^2 == ", round(tail(rf_col_rate$rsq, 1), 2))) # + 
#  theme(legend.position = 'none')

#importance.df <- data.frame(otu = names(importance_final),
#                            importance = importance_final,
#                            get_tax(1, names(importance_final), tax_file))

cr_features_cor <- sapply(as.character(cr_features), 
                       function(i)rownames(rel_abund_cor)[rel_abund_cor[,i] > cor_cutoff])

@@

cage_color <- rainbow(length(unique(cr_prediction$cage_id)))
for (otu in seq_along(cr_features)) {
    partialPlot(rf_col_rate, shared_cr_df, cr_features[otu],
                xlab='', main=paste(get_tax(row_list = cr_features)$tax_label[otu]))#, log = 'x')
    par(new = T)
    plot(y = shared_col_rate_df[ , 'ratio'],
         x = shared_col_rate_df[ , cr_features[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Relative Colonization')
    if(length(cr_features_cor[[otu]]) > 1){
      mtext(side = 1, line = 3, cex = 0.6,
            paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
                  paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ',')))
    }
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(shared_col_rate_df$cage_id)), function(m){
    mouse <- unique(shared_col_rate_df$cage_id)[m]
    points(y = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse, 'ratio'],
           x = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse , cr_features[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```
